---
title: "ECO2411. Assignment 3"
author: "Andriy Levitskyy (ID:1003874701), Yukuo Liu(ID:998402232)"
output: pdf_document
---
The answer to Exercise 1 is after the answer to Exercise 2

# Exercise 2

### 2.2

The ACF of VIX suggest absence of any autocorrelation dynamics for the series. ACF for S&P looks similar to an AR(1) process with small autocorrelation coefficient. Cross-ACF suggest possibility of some Granger-Causality between variables, as the cross-autocorrelations are significant at lags 1,2 and 3.

### 2.4

The most persistant combination is given by the eigenvector on the largest in magnitude eigenvalue of AR(1) coefficient matrix: [1 S&P,0.00063 VIX]
The least risky portfolio can be found by renormalizing the eigenvector on the smalles eigenvalue of the covariance matrix [-1 S&P,0.00013 VIX] in such a way that the sum of entries are equal to 1 (for example by dividing by the sum of the entries). The portfolio shares I got are:
[1.0001300169, -0.0001300169], implying that one needs to short VIX to decrease variance. As the eigenvalue is very close to zero, this may imply that this portfolio is almost riskless.

### 2.6

As squared return on S&P is a good indicator of the volatility of stock market prices, I may decide to increase my exposure to the stock market, as the VAR forecast predicts decrease in risk. Given the results of the spectral decomposition I would also buy some VIX options, so I can be compensated if the stock market volatility increases.
\pagebreak

# References

Kilian (2013). Structural Vector Autoregressions. Handbook of Research Methods and Applications in Empirical Macroeconomics, Chapter 22, 515-554, Edward Elgar.

# Appendix

#### Making the dataset

```{r results='hide', message=FALSE, warning=FALSE}
library(lubridate)
library(zoo)
library(ggplot2)
library(vars)
setwd("/Users/andriylevitskyy/Desktop/MA/Fin Metr/assignment3")
vix<-read.csv("vixcurrent.csv")
vix<-vix[,c(1,2)]
vix$Date<-mdy(vix$Date)
colnames(vix)<-c("date","vix")
sp500<-read.csv("SP500.csv")
sp500$DATE<-ymd(sp500$DATE)
colnames(sp500)<-c("date","sp")
final_dataset<-merge(vix,sp500,by="date")
final_dataset$sp<-as.numeric(as.character(final_dataset$sp))
final_dataset<-read.zoo(final_dataset)
sp2<-diff(log(final_dataset$sp))^2
vix2<-diff(log(final_dataset$vix))
final_dataset<-cbind(final_dataset,sp2)
final_dataset<-cbind(final_dataset,vix2)
final_dataset<- data.frame(index(final_dataset), as.data.frame(final_dataset))
colnames(final_dataset)[1]<-"date"
``` 

#### Plotting the series

```{r}
p1<-ggplot() + 
  geom_line(data = final_dataset, aes(x = date, y = vix2), color = "red")+
  xlab('date') +
  ylab('Return on VIX')
p2<-ggplot() + 
  geom_line(data = final_dataset, aes(x = date, y = sp2), color = "blue") +
  xlab('data') +
  ylab('Squared return on S&P')
require(gridExtra)
grid.arrange(p1,p2, ncol=2)
plot(ts(final_dataset$vix2/1000),col="red",ylab="Y1t/1000, Y2t")
lines(ts(final_dataset$sp2),col="black")
``` 

#### Autocorrelation functions
```{r}
par(mfrow=c(1,2))
acf(final_dataset$sp2,na.action=na.pass,main="ACF S&P")
acf(final_dataset$vix2,na.action=na.pass,main="ACF VIX")

ccf(final_dataset$sp2,final_dataset$vix2,na.action=na.pass,main="CACF")
``` 

#### VAR modelling
```{r}
final_dataset<-na.omit(final_dataset)
var_mod<-VAR(final_dataset[,c(4,5)])
summary(var_mod)
``` 
#### Spectral Decompositions
```{r}
covar_mat<-matrix(data=c(1.396e-08,7.027e-07,7.027e-07,5.323e-03),nrow=2,ncol=2)
coef_mat<-matrix(data=c(2.884e-01,2.347e-04,24.482651,-0.066711 ),nrow=2,ncol=2)
eigen(covar_mat)
eigen(coef_mat)
``` 
#### Predictions
```{r}
prediction<-predict(var_mod,n.ahead=50)
par(mfrow=c(1,2))
plot(prediction$fcst$sp2[,1],type="l",ylab="S&P prediction")
plot(prediction$fcst$vix2[,1],type="l",ylab="VIX prediction")
``` 